---
title: "worksheet01_group2"
output: pdf_document
date: "2023-04-21"
author:
- Samra Hamidovic
- Matanat Mammadli
---

```{r global_options, include = FALSE}
knitr::opts_chunk$set(message = FALSE, 
                      tidy.opts = list(width.cutoff = 70), 
                      tidy = TRUE)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Task 1: t-test

### The t-test assumes normally distributed data. For an increasing number of samples N = {5, 10, 20, 50, 100}, sample from each of these distributions:

- a continuous uniform distribution in the interval [-1, 1]

- a student-t distribution with parameter df = 1

- a discrete uniform distribution with domain {-1, 1}

### For each of these distributions and sample sizes N, sample a large number of times and conduct a two-sided t-test for each sample against a mean of 0 (i.e. Ho: mean is 0). Plot the p-value distributions for the smallest sample size for each distribution. Compare your results to the null distribution, and report how many are below the nominal significance level of alpha = 0.05. For all distributions plot the fraction of significant results as the sample size increases. Discuss what you find

## We start by loading some packages and setting up the variables

```{r}
library(tidyverse)
library(magrittr)
#library(ggplot2)
```

## Defining sample sizes and number of iterations for each sample size
```{r}
N <- c(5, 10, 20, 50, 100)
n_iters <- 10000
```

## Creating data frames to store results
```{r}
results_uniform_continuous <- data.frame(
  distribution = character(),
  sample_size = integer(),
  iteration = integer(),
  p_value = numeric()
)
results_student_t <- data.frame(
  distribution = character(),
  sample_size = integer(),
  iteration = integer(),
  p_value = numeric()
)
results_discrete_uniform <- data.frame(
  distribution = character(),
  sample_size = integer(),
  iteration = integer(),
  p_value = numeric()
)
```

## Looping over sample sizes and distributions, conducting t-tests and storing results

### Continuous uniform distribution on [-1, 1]
```{r}
# Have to loop over each distribution separately because it kept throwing 
# "Error in t.test.default(data) : data are essentially constant"
for (n in N) {
  for (i in 1:n_iters) {
    data <- runif(n, min = -1, max = 1)
    p_value <- t.test(data)$p.value #two-sided by default
    results_uniform_continuous <- results_uniform_continuous %>%
      add_row(
        distribution = "uniform",
        sample_size = n,
        iteration = i,
        p_value = p_value
      )
  }
}
```

### Student-t distribution with degrees of freedom = 1
```{r}
for (n in N) {
  for (i in 1:n_iters) {
    data <- rt(n, df = 1)
    p_value <- t.test(data)$p.value
    results_student_t <- results_student_t %>% add_row(
      distribution = "Student-t",
      sample_size = n,
      iteration = i,
      p_value = p_value
    )
  }
}
```

### Discrete uniform distribution on {-1, 1}
```{r}
set.seed(123)
for (n in N) {
  if (n == 5 || n == 10) {
    n_iters = 30 # only seems to be working when sampled less number of times 
    # for a small sample size else it throws the error 
    # 'data are essentially constant'
  }
  else {
    n_iters = 10000 # works for big enough sample size
  }
  for (i in 1:n_iters) {
    data <- sample(c(-1, 1), n, replace = TRUE)
    p_value <- t.test(data)$p.value
    results_discrete_uniform <- results_discrete_uniform %>% add_row(
      distribution = "Discrete Uniform",
      sample_size = n,
      iteration = i,
      p_value = p_value
    )
  }
}
```

## Plotting the p-value distributions for the smallest sample size of each distribution.
```{r}
results_uniform_continuous %>%
  filter(sample_size == min(N)) %>%
  pull(p_value) %>% # extract the 'p_value; column as a numeric vector
  hist(breaks = 10, main = "p-value distribution(Uniform continuous on [-1, 1])", 
       xlab = "p-values", col = "#8669d355") # plot the histogram
```

p-values are distributed uniformly (as data is uniformly sampled)

```{r}
results_student_t %>%
  filter(sample_size == min(N)) %>%
  pull(p_value) %>% # extract the 'p_value; column as a numeric vector
  hist(breaks = 30, main = "p-value distribution(Student-t with df = 1)", 
       xlab = "p-values", col = "#8669d355") # plot the histogram
```

p-values are a bit skewed to the left.

```{r}
results_discrete_uniform %>%
  filter(sample_size == min(N)) %>%
  pull(p_value) %>% # extract the 'p_value; column as a numeric vector
  hist(breaks = 10, main = "p-value distribution(Uniform discrete on {-1, 1})", 
       xlab = "p-values", col = "#8669d355") # plot the histogram
```

## Calculating the number of significant results i.e. below  the nominal significance level of alpha = 0.05 

### Continuous uniform distribution on [-1, 1]
```{r, message = FALSE, warning = FALSE}
significant_results_continuous <- results_uniform_continuous %>%
  filter(p_value < 0.05) %>%
  group_by(sample_size) %>%
  summarize(num_of_significant_p_values = n()) %>%
  mutate(fraction = num_of_significant_p_values / nrow(results_uniform_continuous))
print(significant_results_continuous)
```

### Student-t distribution with degrees of freedom = 1
```{r, message = FALSE, warning = FALSE}
significant_results_student_t <- results_student_t %>%
  filter(p_value < 0.05) %>%
  group_by(sample_size) %>%
  summarize(num_of_significant_p_values = n()) %>%
  mutate(fraction = num_of_significant_p_values / nrow(results_student_t))
print(significant_results_student_t)
```

### Discrete uniform distribution on {-1, 1}
```{r}
significant_results_discrete <- results_discrete_uniform %>%
  filter(p_value < 0.05) %>%
  group_by(sample_size) %>%
  summarize(num_of_significant_p_values = n()) %>%
  mutate(fraction = num_of_significant_p_values / nrow(results_discrete_uniform))
print(significant_results_discrete)
```

## Plotting the fraction of significant results as the sample size increases

### Continuous uniform distribution on [-1, 1]
```{r}
plot(significant_results_continuous$sample_size, significant_results_continuous$fraction, 
     type = "o", 
     xlab = "sample size", 
     ylab = "fraction of significant results",
     main = "Fraction of Significant Results vs. Sample Size \nContinuous uniform distribution on [-1, 1]",
     col = "#311B92",
     lwd = 2
)
```

### Student-t distribution with degrees of freedom = 1
```{r}
plot(significant_results_student_t$sample_size, significant_results_student_t$fraction, 
     type = "o", 
     xlab = "sample size", 
     ylab = "fraction of significant results",
     main = "Fraction of Significant Results vs. Sample Size \nStudent-t distribution with degrees of freedom = 1",
     col = "#311B92",
     lwd = 2
)
```

### Discrete uniform distribution on {-1, 1}
```{r}
plot(significant_results_discrete$sample_size, significant_results_discrete$fraction, 
     type = "o", 
     xlab = "sample size", 
     ylab = "fraction of significant results",
     main = "Fraction of Significant Results vs. Sample Size \nDiscrete uniform distribution on {-1, 1}",
     col = "#311B92",
     lwd = 2
)
```

### Here, we can observe that as the sample size increases, the fraction of significant results decreases (for the uniform distribution), which is maybe indicative of getting less surprising results (the true mean is closer to our null hypothesis value of 0). Whereas for the student-t an uniform discrete distributions, the fraction of p-values increases, which indicates that the true mean is further away from our null hypothesis of 0.


# Task 2: t-tests

### A small clinical trial on asthma patients has been run measuring the lung function of a control group on a placebo and a treatment group on a new drug.

```{r}
library(tidyverse)
#library(magrittr)
library(ggplot2)
```

##  Reading in the data from the CSV file
```{r}
lung_data <- read_csv('./lung_data.csv')
head(lung_data)
```

## Visualizing the data for each group,

### Boxplot
```{r}
ggplot(lung_data, aes(x = Trial.arm, y = Lung.function, fill = Trial.arm)) +
  geom_boxplot() +
  labs(x = "Group", y = "Lung Function", 
       title = "Boxplot of Lung Function by Group")
```

### Violin plot (used to show the distribution of lung function values for each group, as well as the median and quartile ranges)
```{r}
ggplot(lung_data, aes(x = Trial.arm, y = Lung.function,  fill = Trial.arm)) +
  geom_violin() +
  labs(x = "Group", y = "Lung function",
       title = "Violin plot of Lung function by group")
```

### Density plot (to show the probability density of lung function values for each group)
```{r}
ggplot(lung_data, aes(x = Lung.function, fill = Trial.arm)) +
  geom_density(alpha = 0.5) +
  labs(x = "Lung function", y = "Density",
       title = "Density plot of Lung function by group") +
  scale_fill_manual(values = c("#f8766d", "#00bfc4"), name = "Group")
```

### Bee swarm plot(used to show the distribution of lung function values for each group)
```{r}
ggplot(lung_data, aes(x = Trial.arm, y = Lung.function, color = Trial.arm)) +
  geom_point(position = position_jitter(width = 0.2, height = 0), alpha = 0.5) +
  labs(x = "Group", y = "Lung function", 
       title = "Bee swarm plot of Lung function by group") +
  scale_color_manual(values = c("#f8766d", "#00bfc4"), name = "Group") +
  theme(legend.position = "none")
```

### Q-Q plot
```{r}
# Calculate the theoretical quantiles for a normal distribution
expected_qq <- qnorm(p = seq(0.01, 0.99, by = 0.01))

# Create the QQ plot
ggplot(lung_data, aes(sample = Lung.function)) +
  stat_qq(distribution = qnorm) +
  stat_qq_line() +
  labs(title = "Normal Q-Q Plot of Lung Function") +
  xlab("Theoretical Quantiles") +
  ylab("Sample Quantiles")
```

We can see our data is more or less normally distributed.

## Performing the two-sample t-test to test whether there is a difference in function between the two groups
```{r}
t.test(Lung.function ~ Trial.arm, data = lung_data)
```

### From the t-test, we can see that there is a difference of 0.325 between the means of both the groups. The resultant p-value is 0.0374 which is less than our significant threshold of 0.05. This suggests that it is less likely that the observed difference in means happened by accident, under the null. Hence, we may say that the new medication, statistically, significantly affects lung function when compared to a placebo.
